YALE DEPARTMENT OF ECONOMICS \\ \vspace*{60pt} The Folly of the Crowd: \\ Market Inefficiencies in Peer to Peer Lending

### 2015-03-31 draft.Rnw
Made timeline. Added sentence to conclusion.

\documentclass[12 pt]{article}
%\usepackage[doublespacing]{setspace}
\usepackage{graphicx, verbatim}
\usepackage[font=small,labelfont=sf]{caption}
\usepackage{chronosys}
\usepackage{float}
\usepackage{subfig}
\usepackage{url}
\setlength{\textwidth}{7in} 
\setlength{\textheight}{9in}
\setlength{\oddsidemargin}{0in} 
\setlength{\evensidemargin}{0in}
\setlength{\topmargin}{-1.5cm}
\overfullrule=2cm
\usepackage{titlesec}
\usepackage[linktoc=all]{hyperref}
\hypersetup{colorlinks, citecolor=black, filecolor=black, linkcolor=black,urlcolor=black}
\usepackage[citestyle=verbose,bibstyle=numeric,sorting=ynt,backend=biber]{biblatex}
\usepackage[toc,page]{appendix}
\addbibresource{lit.bib}

\begin{document}
\setcounter{tocdepth}{2}
%\linespread{1.3}
\title{YALE DEPARTMENT OF ECONOMICS \\ \vspace*{60pt} The Folly of the Crowd: \\ Market Inefficiencies in Peer to Peer Lending}
\author{Samuel Shleifer\footnote{R code available at
\url{github.com/sshleifer/ProsperThesis} TODO(THANKS(VK, Pope, AS, Nick, Iyer, Metrick)}}
\date{\today}
\maketitle
\begin{center}
Essay submitted to the Faculty of the Department of Economics in partial fulfillment
of the requirements for the degree of Bachelor of Arts at Yale University.

@@

Prosper.com data provide some additional valuable insights into investor
behavior. Over the period I study, there are three interest rate determination regimes:
\begin{enumerate}
\item Early Auction Regime (April 19, 2006-October 15, 2008): Rates are set by investor auction. Investors have large amounts of data, but Prosper does not guide them meaningfully.
\item Late Auction Regime: (July 1, 2009-December 20, 2010): Rates are set by investor auction, with Prosper offering valuable guidance in the form of more complex credit grading.
\item Prosper-rate Regime: (December 21, 2010 - August 3, 2011): Prosper sets rates using its more complex credit grading algorithm.
\end{enumerate}
Additionally, for regimes (2) and (3), subprime borrowers are not allowed to
apply for Prosper loans.

@@

Much of the evidence presented in this essay is inconsistent with an explanation
that all of the differences between regimes can be explained by differences in
time period. For example, in the late auction regime, when investors set
interest rates, strategies using publicly available information can earn excess
reutrns. In the Prosper-rate regime, when Prosper sets interest rates directly,
they cannot. The two periods are very similar macroeconomically. Perhaps most tellingly,
in the period when Prsper sets rates it seems to use precisely the information that was
helpful in selecting the best performing loans in the early auction regime.
The points strongly toward the conclusion that initial investors were not using
publicly available information.

@@

@@

\iffalse
\begin{table}
  \renewcommand\arraystretch{1.4}\arrayrulecolor{LightSteelBlue3}
  \captionsetup{singlelinecheck=false, font=blue, labelfont=sc, labelsep=quad}
  \caption{Timeline}\vskip -1.5ex
  \begin{tabular}{@{\,}r <{\hskip 2pt} !{\foo}
    >{\raggedright\arraybackslash}p{5cm}}
    \toprule
    \addlinespace[1.5ex]
    1947 & AT and T Bell Labs develop the idea of cellular phones\\
    1968 & Xerox Palo Alto Research Centre envisage the 'Dynabook\\
    1971 & Busicom 'Handy-LE' Calculator\\
    1973 & First mobile handset invented by Martin Cooper\\
    1978 & Parker Bros. Merlin Computer Toy\\
    1981 & Osborne 1 Portable Computer\\
    1982 & Grid Compass 1100 Clamshell Laptop\\
    1983 & TRS-80 Model 100 Portable PC\\
    1984 & Psion Organiser Handheld Computer\\
    1991 & Psion Series 3 Minicomputer\\
  \end{tabular}
\end{table}
\fi
\iffalse
\startchronology
[startyear=2006,stopyear=2011,height=1ex,width=\hsize]
\chronoevent{0}{Jan 13 Quiz 1}
\chronoevent{8}{Jan 20 Quiz 2}
\chronoevent{36}{Feb 17 Quiz 6}
\chronoevent{36}{Intervention}
\stopchronology
\fi
In my sample, Prosper's origination volume, as shown in Figure \ref{fig:origTime}, peaked in the months before the shutdown, and then slowly resumed after the quiet period. Since 2011, when the sample ends, origination volume has eclipsed its 2008 peak.\newline
<<origTime, cache=FALSE, fig.cap='This chart shows two panels. The first panel shows the number of loans funded on Prosper.com in each month of the sample. The second panel shows the total dollars lent to borrowers on Propers.com in each month of the sample.'>>=
g2 = group_by(nd, mgroup, regime) %>%
  summarise(Number_Of_Loans = length(mgroup) 
            ,Dollars_Lent = sum(LoanOriginalAmount)) %>%
    melt(id.vars = c("mgroup",'regime'))

@@

\subsection{Auction Design}
From April 19, 2006 until December 20, 2010. Prosper operated a \emph{variable
rate} model, where lenders and borrowers determined interest rates using a Dutch
auction-like system, where each investor bid consisted of a (dollar amount,
interest rate) pair, such that the dollar amount is greater than \$50, the
minimum bid, and the rate is capped at the borrower's reservation
rate.\footcite{freedman}
In these auctions, borrowers would declare a maximum reservation interest
rate, capped at 36\%, where bidding would start. Over the course of a 7-10 day bidding period, that rate would be bid down by investors, with the loan's interest rate closing at the lowest rate $R_{clearing}$ such that enough investors were willing to fully fund the loan at or below the $R_{clearing}$. All investors who bid at or below the clearing rate would receive the clearing rate. If a borrower could not reach 100\% funding at his reservation rate, the loan would not be funded.\newline
As an instructional example, suppose Bill the borrower needed to borrow \$100, and set his maximum rate at 20\%.
Let Chris, David and Ellen be investors, with Chris bidding (\$50, 15\%), David bidding (\$50, 18\%) and Ellen bidding (\$30,19\%). In this scenario, Chris and David would win the auction, and each lend lend Bill \$50, and receive an 18\% interest rate. Bill would pay 18\% interest before fees.\newline

\subsection{Early Auction Regime}
In the early auction regime, loans originated between April 19, 2006 and October 15, 2008, auction participants had access to considerable information from Prosper, including over 60  Borrower characteristics, such as residence, employment status and a FICO score bucket, but few intelligent risk estimates. Additionally, Prosper did not regulate who could apply for a loan on the website. Besides basic fraud prevention, any applicant who received 100\% funding from investors was eligible, for Prosper's first year of operation. On March 1, 2007, amidst poor returns, Prosper announced that listings with credit scores below 520 would no longer be offered to investors. This intervention is not discussed for lack of space, but shows that after a year, more or less, Prosper felt the need to nudge a fully decentralized investment process in the right direction.\newline
In an  October 15, 2008 email, Prosper announced that it would temporarily stop accepting new loan applications to register with the SEC, which subsequently released a cease and desist order characterizing Prosper as a ``seller of securities," rather than a marketplace merely matching lenders and borrowers, as the site had argued.
\footnote{\url{techcrunch.com/2008/11/26/sec-outlines-its-reasoning-for-shutting-down-p2p-lender-prosper/}}\footnote{SEC LETTER}\newline
\subsection{Late Auction Regime}
After a 9 month quiet period, Prosper re-opened, using the same Auction model,
with three significant market interventions. First, under-writing standards
again became more stringent: new borrowers were required to have a credit score score above 640, and repeat borrowers were required to have a score above 600.

Second, whereas previously Prosper used only the credit score to determine a borrower's Credit Grade (eg. AA), the site began offering a more advanced, multi variable formula to determine each borrower's credit grade for investors consideration. Finally, the site also partnered with FolioFN, a secondary market where lenders could liquidate their positions. Crucially, Prosper still had no direct role in setting interest rates, investors still used an auction, but with the extra information credit grades that more closely matched default probabilities. 
Besides credit score, Prosper began using (and still uses) the following ``Key
variables" to determine its revamped credit grades:\footcite{estloss}
\begin{itemize}
\item  TotalInquiries: The number of times potential lenders requested the borrowers' credit portfolio from the credit bureau, which is usually Experian or a competitor.
\item CurrentDelinquencies: The number of delinquent loans associated with the borrower.
\item Credit card utilization: how much of available credit the borrower uses on a monthly basis.
\item Number of open credit card accounts
\item Debt to income ratio
\item Loan payment performance on prior Prosper loans.
\end{itemize}
This Auction rate-setting mechanism continued for more than a year after the shutdown, until December 20, 2010.

Unsuprisingly, buut critically for the interpretation of the evidence,
the automated feature selection procedure in section 5, which
identifies the subset of variables that best predict default, finds that
credit Score, inquiries, delinquencies and number of credit cards are all useful
independent variables for predicting default.

\subsection{Prosper-rate Regime}
Under the current \emph{Prosper-rate} model, enacted December 21, 2010, lenders
no longer determine the loan rate in an auction. Instead, Prosper uses the
intelligent credit grades given to investors in the late auction regime to set
interest rates on loans, with slightly less stringent under-writing standards to
determine eligibility. The change was motivated by a desire to accelerate the
listing process, so that investors could reinvest money more quickly. Moreover,
towards the end of the late auction regime, auctions were arriving at interest
rates that were almost completely determined by the credit grade. Under the
fixed-rate regime, investors choose whether or not to invest in a given loan, at
the interest rate set by Prosper, so bids consist only of a dollar amount.\newline
%TODO: NEW PAGE IF CRUNCHED
\section{Literature}
\subsection{Efficient Markets Hypothesis}
In layman's terms, the efficient markets hypothesis (EMH), states that 'it is
impossible to beat the market.' More specifically, Fama explains that market
investors (or accepted by investors in the fixed-rate regime) at origination
time, and paid by the borrower, and returns are the cash flows received from borrowers divided by the amount invested.
Fama, (1970) notes the dependence of most tests of market efficiency on a model of risk and return.
The models, including most famously, Sharpe's capital asset pricing model ('CAPM'), make two predictions:
First,  since an investor can always hold cash risklessly, the average return on risky investments must be positive. Second, securities with higher systematic risk earn higher average returns.\footcite{sharpe}

@@

\subsection{Measuring Loan Performance with ROI}
<<roivstatus,cache=FALSE,fig.height=3,fig.cap="This chart shows the distribution of ROI for completed, defaulted and chargedoff loans over all 3 regimes. A continuous ROI variable offers more information about the quality of an investment than a binary default indicator. Dotted lines indicate group means.",fig.height=3>>=
smeans = summarise(group_by(nd,LoanStatus), mRoi = mean(roi, na.rm=TRUE))
ggplot(nd, aes(x=roi,fill=LoanStatus,scales='free_y')) +
  geom_histogram(binwidth=.01)+
  facet_wrap(~LoanStatus,ncol=1) +
  geom_vline(data=smeans, aes(xintercept=mRoi,color=LoanStatus),linetype='longdash') + 
  theme_bw() +
  # facet_wrap(~LoanStatus,ncol = 1)+
  labs(title='Distribution of ROI for each Loan Outcome (all regimes)', x='ROI',y='Count') +
  scale_x_continuous(limits=c(-1, 1), breaks=seq(-1,1,by=.25),label=percent)
  # scale_y_continuous(label=comma)
  @

@@

I do not discount or annualize ROIs, because in the period of low interest
rates, the difference between default and repayment is large enough that a few
percentage points don't matter: loans that default average -46\% ROI, loans that
repay average + 24\% ROI. Furthermore, my ROI results are very similar to Freedman
and Jin's, who use discounting.\footcite{freedman} %TODO(SS): CITE FED RATES

@@

For credit buckets above 600, the relationship between credit score and default rate appears to be fairly linear, for all three regimes, suggesting that borrowers should be sensitive to credit scores when setting interest rates, since they are predictive of default.\footnote{I have a chart (suggested by Lapo) that shows what rate investors would have needed to charge to break even. It shows, as expected, that investors in early regime would have needed to charge much higher rates, especially at risky credit grades to break even. Should I put it in?}
Early auction regime investors, then, were not sufficiently sensitive to the
hard credit variables that are predictive of default, especially when evaluating
risky, low-credit borrowers, suggesting that the poor returns of the early
auction regime were not caused exclusively by the time period, but also by poor investor decisions. We now look at the more informative grading algorithm Prosper began using after the shutdown, which made investors more sensitive to hard credit variables.

\subsubsection {More Informative Grading, Higher Interest Rates}
<<csmix2, cache=FALSE, fig.height=9, fig.cap='Another way to visualize the credit scores of borrowers in each grade. Dotted Lines indicate mean credit bucket for each grade. Each (row,column) entry is the distribution of borrower credit scores that were assigned grade (column) during the (row) regime',fig.height=5>>=
gbf = filter(rbind(o1,o2), grade != 'NC') %>% filter(grade != '') %>% group_by(grade,regime)
gbf$g2 = factor(gbf$grade, levels= c("AA","A","B","C","D","E","HR"))
means  = summarise(gbf, val = mean(CreditScoreEst,na.rm = TRUE)
                   ,num = length(CreditScoreEst)) %>% transform(g2= grade)

@@

In the late auction regime, Prosper began using multiple variables to form its credit grade, and borrowers in each credit score bucket are spread across grades based on other credit variables. Correspondingly, in the bottom row of the grade C column of Figure \ref{fig:csmix2}, we can see that late auction regime, grade C borrowers came from 11 different credit buckets.
When Prosper began aggregating hard credit variables into more informative grades, investors relying heavily on these grades when determining interest rates.
<<ratepred,cache=FALSE>>=
func2 <- function(LenderYield,CreditScoreEst,TotalInquiries,CurrentDelinquencies,grade){
  tmp = data.frame(cbind(LenderYield,CreditScoreEst,TotalInquiries,CurrentDelinquencies,factor(grade)))
  rateReg  = mlm(LenderYield ~ CreditScoreEst + CurrentDelinquencies + grade, data=tmp)
  return(rateReg$adj.r.squared)
}
func3 <- function(LenderYield,CreditScoreEst,TotalInquiries,CurrentDelinquencies,grade){
  tmp = data.frame(cbind(LenderYield,CreditScoreEst,TotalInquiries,CurrentDelinquencies,factor(grade)))
  rateReg  = mlm(LenderYield ~ CreditScoreEst + CurrentDelinquencies, data=tmp)
  return(rateReg$adj.r.squared)
}
func4 <- function(LenderYield,CreditScoreEst,TotalInquiries,CurrentDelinquencies,grade,defaulted){
  tmp = data.frame(cbind(LenderYield,CreditScoreEst,TotalInquiries,CurrentDelinquencies,factor(grade)))
  rateReg  = mlm(defaulted~ CreditScoreEst, data=tmp)
  return(rateReg$adj.r.squared)
}

@@

rf2 = group_by(nd,regime) %>% 
  summarise(fit = func3(LenderYield,CreditScoreEst,TotalInquiries,CurrentDelinquencies,grade)
            ,num = length(LenderYield))
colnames(rf2) = c('Regime','Fit','Number of Loans')
xtable(rf2, label='ratefit2',caption= 'This table documents how much of the variance in interest rates can be explained by borrowers\' credit score buckets. \'Fit\' is measured as the Adjusted $R^{2}$ with interest rate as the dependent variable and credit score bucket as the independent variable, using every loan in the relevant regime.')
@
The tension between the importance of the credit score in the credit grade and the importance of the credit score in the interest rate is resolved by the fact that credit grades are much more predictive of interest rates in the later two regimes. In Table \ref{ratefit}, the \emph{Fit} column shows the fraction of the variance in interest rate that can be described by variance in hard credit variables, as measured by the adjusted $R^{2}$ of a regression with interest rate as the dependent variable and credit score bucket, current delinquencies and credit grade as independent variables. This regression yields adjusted $R^{2}$ of 48\%, 88\% and 97\% for the early auction, late auction and Prosper-rate regimes, respectively, suggesting that Prosper's more informative grades are immediately used by borrowers to determine interest rates. \footnote{The full regression results are available in the appendix.}

Since we do not know Propser's exact grading formula, it may include information
that was not previously available to borrowers. Table \ref{ratefit2} addresses this concern, by running the same procedure as Table \ref{ratefit}, but with only credit score bucket as an independent variable. The direction of the results is the same: Prosper's change to more informative grading did not only lead to a larger role for those grades in investors' interest rates, but increased the weight of credit score bucket, a variable that investors had access to in the same format in both periods. If the crowd of investors were as wise as Hayek hopes, Prosper's repackaging of multiple variables into a more informative credit grade should not affect the weight of hard information in investors' risk estimates. Furthermore, Figure \ref{fig:defvscore} makes it clear that investors would have experienced fewer defaults, and thus earned higher returns, if they had used credit buckets more heavily in their risk assessments, given the poor returns associated with low credit borrowers. We can thus attribute some of the poor returns of the early auction regime to a lower level of reliance on hard information.

>>>>

\documentclass[12 pt, twoside]{article}
%\usepackage[doublespacing]{setspace}
\usepackage{graphicx, verbatim}
\usepackage[twoside, bindingoffset=5mm]{geometry}
\usepackage[font=small,labelfont=sf]{caption}
%\usepackage{chronosys}
\usepackage{float}
\usepackage{subfig}
\usepackage{url}
%%TIMELINE SHIT
\usepackage{array, booktabs} % for timeline
\usepackage[x11names]{xcolor}
\usepackage{colortbl}
\usepackage{caption}
\DeclareCaptionFont{blue}{\color{LightSteelBlue3}}

\newcommand{\foo}{\color{LightSteelBlue3}\makebox[0pt]{\textbullet}\hskip-0.5pt\vrule width 1pt\hspace{\labelsep}}
%%END
\setlength{\textwidth}{7in} 
\setlength{\textheight}{9in}
\setlength{\oddsidemargin}{0in} 
\setlength{\evensidemargin}{0in}
\setlength{\topmargin}{-1.5cm}
\overfullrule=2cm
\usepackage{titlesec}
\usepackage[linktoc=all]{hyperref}
\hypersetup{colorlinks, citecolor=black, filecolor=black, linkcolor=black,urlcolor=black}
\usepackage[citestyle=verbose,bibstyle=numeric,sorting=ynt,backend=biber]{biblatex}
\usepackage[toc,page]{appendix}
\addbibresource{lit.bib}

\begin{document}
\setcounter{tocdepth}{2}
%\linespread{1.3}
\title{YALE DEPARTMENT OF ECONOMICS \\ \vspace*{60pt} The Folly of the Crowd: \\ Market Inefficiencies in Peer to Peer Lending}
\author{Samuel Shleifer
  \footnote{Thanks to Professor Vineet Kumar for his help
  in developing the ideas and statistical methodologies involved in this paper,
  to Devin Pope and Rajkamal Iyer for help manipulating the Prosper data, 
  and to Andrei Shleifer, Nicholas Barberis and Andrew Metrick for helpful comments and suggestions.}
  \footnote{R code available at \url{github.com/sshleifer/ProsperThesis}}}
\date{\today}
\maketitle
\begin{center}
Essay submitted to the Faculty of the Department of Economics in partial fulfillment
of the requirements for the degree of Bachelor of Arts at Yale University.

@@

Prosper.com data provide some additional valuable insights into investor
behavior. Over the period I study, there are three interest rate determination regimes:
\begin{enumerate}
\item Early Auction Regime (April 19, 2006-October 15, 2008): Rates are set by investor auction. Investors have large amounts of data, but Prosper does not guide them meaningfully.
\item Late Auction Regime: (July 20, 2009-December 19, 2010): Rates are set by investor auction, with Prosper offering valuable guidance in the form of more complex credit grading.
\item Prosper-rate Regime: (December 20, 2010 - August 3, 2011): Prosper sets rates using its more complex credit grading algorithm.
\end{enumerate}
Additionally, for regimes (2) and (3), subprime borrowers are not allowed to
apply for Prosper loans.

@@

Much of the evidence presented in this essay is inconsistent with an explanation
that all of the differences between regimes can be explained by differences in
time period. For example, in the late auction regime, when investors set
interest rates, strategies using publicly available information can earn excess
returns. In the Prosper-rate regime, when Prosper sets interest rates directly,
they cannot. The two periods are very similar macroeconomically. Perhaps most tellingly,
in the period when Prosper sets rates it seems to use precisely the information that was
helpful in selecting the best performing loans in the early auction regime.
The points strongly toward the conclusion that early investors were not using
publicly available information effectively.

@@

\begin{table}
  %\renewcommand\arraystretch{1.4}\arrayrulecolor{LightSteelBlue3}
  %\captionsetup{singlelinecheck=false, font=blue, labelfont=sc, labelsep=quad}
  \caption{Timeline of Events}\vskip -1.5ex
  \begin{tabular}{@{\,}r <{\hskip 2pt} !{\foo}
    >{\raggedright\arraybackslash}p{12cm}}
    \toprule
    \addlinespace[1.5ex]
    Apr. 21, 2006 & First loan of early auction regime.\\
    Mar. 1, 2007 & Prosper bans borrowers with credit score $\leq$ 520.\\
    Oct. 15, 2008 & Early Auction period ends with SEC enforced shutdown.\\
    Jul. 20, 2009 & Prosper reopens with more informative credit grades, requires repeat borrowers to have credit score
    $\geq$ 600, and requires new borrowers to have credit score $\geq$ 640.
    Beginning of Early Auction Regime.\\
    Dec. 19, 2010 & Prosper-rate regime, where Prosper sets rates directly,
    begins, because "auctions made the deployment of funds more time consuming
    with little gain in lender returns,” - Prosper CEO Chris
    Larsen.\footcite{renton}\\
    Aug. 3, 2011 & Last loan in Prosper-rate regime originated.
  \end{tabular}
\end{table}[h]

@@

In my sample, Prosper's origination volume, as shown in Figure \ref{fig:origTime}, peaked in the months before the shutdown, and then slowly resumed after the quiet period. Since 2011, when the sample ends, origination volume has eclipsed its 2008 peak.\newline
<<origTime, cache=FALSE, fig.cap='This chart shows two panels. The first panel shows the number of loans funded on Prosper.com in each month of the sample. The second panel shows the total dollars lent to borrowers on Propers.com in each month of the sample.'>>=
g2 = group_by(nd, mgroup, regime) %>%
  summarise(Number_Of_Loans = length(mgroup) 
            ,Dollars_Lent = sum(LoanOriginalAmount)) %>%
    melt(id.vars = c("mgroup",'regime'))

@@

\subsection{Auction Design}
From April 19, 2006 until December 19, 2010. Prosper operated a \emph{variable
rate} model, where lenders and borrowers determined interest rates using a Dutch
auction-like system, where each investor bid consisted of a (dollar amount,
interest rate) pair, such that the dollar amount is greater than \$50, the
minimum bid, and the rate is capped at the borrower's reservation
rate.\footcite{freedman}
In these auctions, borrowers would declare a maximum reservation interest
rate, capped at 36\%, where bidding would start. Over the course of a 7-10 day bidding period, that rate would be bid down by investors, with the loan's interest rate closing at the lowest rate $R_{clearing}$ such that enough investors were willing to fully fund the loan at or below the $R_{clearing}$. All investors who bid at or below the clearing rate would receive the clearing rate. If a borrower could not reach 100\% funding at his reservation rate, the loan would not be funded.\newline
As an instructional example, suppose Bill the borrower needed to borrow \$100, and set his maximum rate at 20\%.
Let Chris, David and Ellen be investors, with Chris bidding (\$50, 15\%), David bidding (\$50, 18\%) and Ellen bidding (\$30,19\%). In this scenario, Chris and David would win the auction, and each lend lend Bill \$50, and receive an 18\% interest rate. Bill would pay 18\% interest before fees.\newline

\subsection{Early Auction Regime}
In the early auction regime, loans originated between April 19, 2006 and October
15, 2008, auction participants had access to considerable information from
Prosper, including over 60  Borrower characteristics, such as residence,
employment status and an Experian credit score bucket, but few intelligent risk estimates. Additionally, Prosper did not regulate who could apply for a loan on the website. Besides basic fraud prevention, any applicant who received 100\% funding from investors was eligible, for Prosper's first year of operation. On March 1, 2007, amidst poor returns, Prosper announced that listings with credit scores below 520 would no longer be offered to investors. This intervention is not discussed for lack of space, but shows that after a year, more or less, Prosper felt the need to nudge a fully decentralized investment process in the right direction.

In an  October 15, 2008 email, Prosper announced that it would temporarily stop accepting new loan applications to register with the SEC, which subsequently released a cease and desist order characterizing Prosper as a ``seller of securities," rather than a marketplace merely matching lenders and borrowers, as the site had argued.\footcite{sec}

\subsection{Late Auction Regime}
After a 9 month quiet period, Prosper re-opened, using the same Auction model,
with three significant market interventions. First, under-writing standards
again became more stringent: new borrowers were required to have a credit score score above 640, and repeat borrowers were required to have a score above 600.

Second, whereas previously Prosper used only the credit score to determine a borrower's Credit Grade (eg. AA), the site began offering a more advanced, multi variable formula to determine each borrower's credit grade for investors consideration. Finally, the site also partnered with FolioFN, a secondary market where lenders could liquidate their positions. Crucially, Prosper still had no direct role in setting interest rates, investors still used an auction, but with the extra information credit grades that more closely matched default probabilities. 
Besides credit score, Prosper began using (and still uses) the following ``Key
variables" to determine its revamped credit grades:\footcite{estloss}
\begin{itemize}
\item  TotalInquiries: The number of times potential lenders requested the borrowers' credit portfolio from the credit bureau, which is usually Experian or a competitor.
\item CurrentDelinquencies: The number of delinquent loans associated with the borrower.
\item Credit card utilization: how much of available credit the borrower uses on a monthly basis.
\item Number of open credit card accounts
\item Debt to income ratio
\item Loan payment performance on prior Prosper loans.
\end{itemize}
This Auction rate-setting mechanism continued for more than a year after the shutdown, until December 19, 2010.

Unsurprisingly, but critically for the interpretation of the evidence,
the automated feature selection procedure in section 5, which
identifies the subset of variables that best predict default, finds that
credit Score, inquiries, delinquencies and number of credit cards are all useful
independent variables for predicting default.

\subsection{Prosper-rate Regime}
Under the current \emph{Prosper-rate} model, enacted December 20, 2010, lenders
no longer determine the loan rate in an auction. Instead, Prosper uses the
intelligent credit grades given to investors in the late auction regime to set
interest rates on loans, with slightly less stringent under-writing standards to
determine eligibility. The change was motivated by a desire to accelerate the
listing process, so that investors could reinvest money more quickly. Moreover,
towards the end of the late auction regime, auctions were arriving at interest
rates that were almost completely determined by the credit grade. Under the
fixed-rate regime, investors choose whether or not to invest in a given loan, at
the interest rate set by Prosper, so bids consist only of a dollar amount.\newline

\section{Literature}
\subsection{Efficient Markets Hypothesis}
In layman's terms, the efficient markets hypothesis (EMH), states that 'it is
impossible to beat the market.' More specifically, Fama explains, market prices should always fully reflect publicly available information. Therefore, EMH rules out the possibility of strategies that earn excess returns using only publicly available information. \footcite{fama}
In the P2P lending context, this market price is the interest rate determined by investors (or accepted by investors in the fixed-rate regime), and paid by the borrower, and returns are the cash flows received from borrowers divided by the amount invested.

Fama, (1970) notes the dependence of most tests of market efficiency on a model of risk and return.
The models, including most famously, Sharpe's capital asset pricing model ('CAPM'), make two predictions:
First,  since an investor can always hold cash risklessly, the average return on risky investments must be positive. Second, securities with higher systematic risk earn higher average returns.\footcite{sharpe}

@@

\subsection{Measuring Loan Performance with ROI}
<<roivstatus,cache=FALSE,fig.height=3,fig.cap="This chart shows the distribution of ROI for completed and defaulted loans over all three regimes. A continuous ROI variable offers more information about the quality of an investment than a binary default indicator. Dotted lines indicate mean ROI for category.",fig.height=3>>=
nd$d.factor =  ifelse(nd$defaulted == 1, 'Defaulted','Completed')
smeans = summarise(group_by(nd,d.factor), mRoi = mean(roi, na.rm=TRUE))
ggplot(nd, aes(x=roi,fill=d.factor)) +
  geom_histogram(binwidth=.01)+ 
  geom_vline(data=smeans, aes(xintercept=mRoi,color=d.factor),linetype='longdash') + 
  theme_bw() +
  labs(title='Distribution of ROI for each Loan Outcome (all regimes)', x='ROI',y='Count',fill='Loan Status') +
  scale_x_continuous(limits=c(-1, 1), breaks=seq(-1,1,by=.25),label=percent) +
  facet_wrap(~d.factor,ncol=1,scales='free_y')
  @

@@

I do not discount or annualize ROIs, because in the period of low interest
rates, the difference between default and repayment is large enough that a few
percentage points don't matter: loans that default average -46\% ROI, loans that
repay average + 24\% ROI. Furthermore, my ROI results are very similar to Freedman
and Jin's, who use discounting.\footcite{freedman} Although their sample is
shorter in length, using discounted ROIs for our sample would only make the
results more extreme, as risk-free interest rates were highest during the early
auction regime.\footcite{fred} \footnote{The fed-funds rate's peak during our sample, is at 5\%, during the Early auction regime. It then falls to near 0 for the duration of the later two
  regimes. Discounting would thus lead to significantly lower early auction regime
 ROIs, but only slightly lower ROIs in the later two periods.}

@@

For credit buckets above 600, the relationship between credit score and default rate appears to be fairly linear, for all three regimes, suggesting that borrowers should be sensitive to credit scores when setting interest rates, since they are predictive of default. 
Early auction regime investors, then, were not sufficiently sensitive to the
hard credit variables that are predictive of default, especially when evaluating
risky, low-credit borrowers, suggesting that the poor returns of the early
auction regime were not caused exclusively by the time period, but also by poor investor decisions. We now look at the more informative grading algorithm Prosper began using after the shutdown, which made investors more sensitive to hard credit variables.

\subsubsection {More Informative Grading, Higher Interest Rates}
<<csmix2, cache=FALSE, fig.height=9, fig.cap='Another way to visualize the credit scores of borrowers in each grade. Dotted Lines indicate mean credit bucket for each grade. Each (row,column) entry is the distribution of borrower credit scores that were assigned grade (column) during the (row) regime.',fig.height=5>>=
nd$g2 = factor(nd$grade, levels= c("AA","A","B","C","D","E","HR"))
gbf = filter(nd, grade != 'NC' & Date < dis1) %>% filter(grade != '') %>% group_by(g2,regime)

means  = summarise(gbf, val = mean(CreditScoreEst,na.rm = TRUE)
                   ,num = length(CreditScoreEst))

@@

In the late auction regime, Prosper began using multiple variables to form its credit grade, and borrowers in each credit score bucket are spread across grades based on other credit variables. Correspondingly, in the bottom row of the grade C column of Figure \ref{fig:csmix2}, we can see that late auction regime, grade C borrowers came from 11 different credit buckets.
When Prosper began aggregating hard credit variables into more informative grades, investors relying heavily on these grades when determining interest rates.
<<ratepred,cache=FALSE>>=
func2 <- function(LenderYield,CreditScoreEst,TotalInquiries,CurrentDelinquencies,grade){
  tmp = data.frame(cbind(LenderYield,CreditScoreEst,TotalInquiries,CurrentDelinquencies,factor(grade)))
  rateReg  = mlm(LenderYield ~ CreditScoreEst + CurrentDelinquencies + grade + TotalInquiries, data=tmp)
  return(rateReg$adj.r.squared)
}
func3 <- function(LenderYield,CreditScoreEst,TotalInquiries,CurrentDelinquencies,grade){
  tmp = data.frame(cbind(LenderYield,CreditScoreEst,TotalInquiries,CurrentDelinquencies,factor(grade)))
  rateReg  = mlm(LenderYield ~ CreditScoreEst, data=tmp)
  return(rateReg$adj.r.squared)
}

@@

rf2 = group_by(nd,regime) %>% 
  summarise(fit = func3(LenderYield,CreditScoreEst,TotalInquiries,CurrentDelinquencies,grade)
            ,num = length(LenderYield))
colnames(rf2) = c('Regime','Fit','Number of Loans')
xtable(rf2, label='ratefit2',caption= 'This table documents how much of the variance in interest rates can be explained by borrowers\' credit score buckets. \'Fit\' is measured as the Adjusted $R^{2}$ with interest rate as the dependent variable and credit score bucket as the independent variable, using every loan in the relevant regime.')
@
The tension between the importance of the credit score in the credit grade and the importance of the credit score in the interest rate is resolved by the fact that credit grades are much more predictive of interest rates in the later two regimes. In Table \ref{ratefit}, the \emph{Fit} column shows the fraction of the variance in interest rate that can be described by variance in hard credit variables, as measured by the adjusted $R^{2}$ of a regression with interest rate as the dependent variable and credit score bucket, current delinquencies and credit grade as independent variables. This regression yields adjusted $R^{2}$ of 49\%, 88\% and 97\% for the early auction, late auction and Prosper-rate regimes, respectively, suggesting that Prosper's more informative grades are immediately used by borrowers to determine interest rates.\footnote{The full regression results are available in the appendix.}

Since we do not know Propser's exact grading formula, it may include information
that was not previously available to borrowers. Table \ref{ratefit2} addresses
this concern by running the same procedure as used for Table \ref{ratefit}, but
with credit score bucket as the only independent variable. The direction of the
results is the same: Prosper's change to more informative grading did not only
lead to a larger role for those grades in investors' interest rates, but
increased the weight of credit score bucket, a variable that investors had
access to in the same format in both periods. The difference between the early and late auction regimes' Fit is much larger in Table \ref{ratefit} than in Table \ref{ratefit2}, which suggests that part of the increased predictability of interest rates in the Late Auction regime comes from hard credit variables besides the credit score bucket.

If the crowd of investors were as wise as Hayek hopes, Prosper's repackaging of multiple variables into a more informative credit grade should not affect the weight of hard information in investors' risk estimates. Furthermore, Figure \ref{fig:defvscore} makes it clear that investors would have experienced fewer defaults, and thus earned higher returns, if they had used credit buckets more heavily in their risk assessments, given the poor returns associated with low credit borrowers. We can thus attribute some of the poor returns of the early auction regime to a lower level of reliance on hard information.

In the next sub section, I estimate the impact of Prosper's other interventions, which ban low-credit borrowers.
\subsection{Impact of Subprime Borrower Ban}
<<csvtime,cache=FALSE,fig.height=3, fig.cap="This figure presents average estimated credit score of loans originated each month. The size of each dot represents the number of loans originated each month. The average credit score (using bucket midpoint) increases from 653 in the early auction regime to 714 in the late auction regime, before returning to 705 in the Prosper-rate regime. Grey clouds indicate one standard deviation.">>=
ggplot(mmeans, aes(x=mgroup, y=mcs, color=regime)) +
  geom_point(aes(size=numLoans))+
  geom_smooth(aes(ymax=mcs.up, ymin=mcs.down),stat='identity') +
  geom_line() +
  theme_bw() +
  labs(title='Borrower Credit Scores Over Time', size='Loans Originated',x='Origination Month', y='Mean Borrower Credit Score')
@
We now turn to Prosper's policy changes, which restrict the types of borrowers who investors are allowed to lend to. An auction perspective predicts that the crowd of lenders should be able to effectively screen all candidates, not just low risk candidates. On multiple occasions, however, Prosper has intervened to restrict the borrower pool to high credit borrowers, and thereby improved investor returns. Figure \ref{fig:csvtime} shows the mean borrower credit score for each month in the sample, with the grey clouds showing one standard deviation bounds. Overall, for the three samples in full, the average borrower credit scores increases from 653.2 before the shutdown to 714.2 after the shutdown, then dips to 705 for the Prosper-rate period.
For the first 10 months of Prosper's history, investors choose to lend to borrowers with an average estimated credit score of 609. Prosper's policy change, on March 1, 2007, that prevented loans to ``low-credit borrowers," defined as those with credit score below 520, showed an immediate positive impact on investor decision making and returns.
